# Shapefiles

Shapefiles contain geospatial vector data, also called geometries.
A shapefile has the extension _.shp_ and it is usually accompanied by other files.
Common supplementary files are:

  - _.shx_, which contains the index of the geometries in the data model.
  - _.dbf_, which contains the attributes of the geometries in a table format.
  - _.prj_, which contains the WKT representation of the coordinate reference system.

The core concept of shapefiles is that geographic features are stored using three fundamental geometry types:

  1. Points, defined by coordinates $(x, y)$.
  2. Lines, defined by two points and a line interpolating them.
  3. Polygons, defiend by several lines.

These three fundamental types can also be "stacked", e.g. one spatial feature can be represented by multiple points, lines, and polygons.
In this case, we talk of _multipoints_, _multilines_, and _multipolygons_.
The difference between multilines and polygons is that a polygon inscribe an area of space, that is it creates an "inside" area and an "outside" one, while multilines do not.

## Geometries in R

Let's see an exampled in R using the package _terra_. 
Remember to load the package if you did not already: `library(terra)`.
In _terra_, all the hustle that came with older GIS software has been removed, and geometries can be simply be created using the `vect()` function.

```{r}
xy <- matrix(c(0, 0), ncol = 2)
poi <- vect(xy)
geomtype(poi)
```

This creates our first geometry, a point at the origin.
To be meaningful, however, we must assign a coordinate reference system to our geometry.
This can be done during the initialization itself.

```{r}
poi <- vect(xy, crs = "EPSG:4326")
poi
```

_terra_ DOES IT USE THE MOST FUNDAMENTAL TYPE?


Geometries can be converted into other types by _casting_.
In _terra_, again, this has been made extremely easy.
Let's create some points.

```{r}
xy <- matrix(seq_len(10), ncol = 2)
pois <- vect(xy, crs = "EPSG:4326")
geomtype(pois)
```

And cast them to lines using `as.lines()`.

```{r}
ls <- as.lines(pois)
geomtype(ls)
```

We may want to cast this to polygons, notice however that these lines do not inscribe an area of space, i.e. they are not "closed".

```{r}
plot(ls)
points(pois, cex = 2)
```

Trying to cast these lines as polygons, will return an empty geometry.

```{r}
pol = as.polygons(ls)
geomtype(pol)
```

To cast lines to polygons, two points must be identical.

```{r}
xy <- matrix(seq_len(10), ncol = 2)
xy <- rbind(xy, cbind(5, 6)) #add bottom-right point
xy <- rbind(xy, xy[1, ]) #add again first point
pois <- vect(xy, crs = "EPSG:4326")
ls <- as.lines(pois)
pol <- as.polygons(ls)
pol
```

If you want to cast points to polygons, it is considered best practice to first cast them to lines and then cast these into polygons (as I just did).
Trying to cast points to polygons directly can introduce errors, if it works at all.

## Geometries in Python


```{python}
from shapely.geometry import Point, MultiPoint

poi = Point(0, 0)
poi.wkt
```

```{python}
pois = MultiPoint([[0, 0], [0, 1], [1, 1], [1, 0], [0, 0]])
pois.wkt
```

```{python}
from shapely.geometry import LineString, LinearRing

ls = LineString([[0, 0], [0, 1], [1, 1], [1, 0], [0, 0]])
lr = LinearRing([[0, 0], [0, 1], [1, 1], [1, 0], [0, 0]])
ls.wkt
```

```{python}
from shapely.geometry import Polygon, MultiPolygon

pol = Polygon([[0, 0], [0, 1], [1, 1], [1, 0], [0, 0]])
pols = MultiPolygon([pol for i in range(3)])
pols.wkt
```

Finally, _shapely_ has the `GeometryCollection` type to store geometries of more than one type.

```{python}
from shapely.geometry import GeometryCollection

coll = GeometryCollection([pois, ls, pol])
coll.wkt
```