# Overview of the data we will use

There are some operations that are the staple of GIS.
For example: What is the distance between two features? What is the area of a polygon? What is the value of a raster at specific locations?
To illustrate this with practical examples, we will use some data that you can find in the `data/` folder in the repo <https://github.com/emilio-berti/gis-course>.

```{bash, eval=FALSE, echo=FALSE}
if [[ ! -d "data" ]]
then
  mkdir data #create data directory
else
  echo "data directory already exists"
fi

cd data #move to data

url="https://gisco-services.ec.europa.eu/distribution/v2/countries/download/ref-countries-2020-10m.shp.zip"

wget $url

# unzip files
files=$(ls)
for f in $files
do
  unzip $f
done

unzip CNTR_RG_10M_2020_4326.shp.zip

cd .. #go back to parent directory
```

```{r, eval=FALSE, echo=FALSE}
library(terra)
v <- vect("data/CNTR_RG_10M_2020_4326.shp")
v <- v[v[["EU_STAT"]] == "T", ]
e <- ext(-11.91, 38.2, 33.89, 73.03)
v <- crop(v, e)
writeVector(v, "EU.shp")
```

```{bash, eval=FALSE, echo=FALSE}
cd data
rm *
mv ../EU.* .
mkdir EU
mv EU.* EU
cd ..
```

`data/EU/EU.shp` contains the polygons of the countries in the European Union:
```{r eu-show}
eu <- vect("data/EU/EU.shp")
plot(eu, col = "#001489", border = "white", lw = 2)
```

`data/ISTAT/Limiti01012024_g/Reg01012024_g/Reg01012024_g_WGS84.shp` contains the polygons of the regions (administrative division below the state) of Italy:
```{r reg-show}
reg <- vect("data/ISTAT/Limiti01012024_g/Reg01012024_g/Reg01012024_g_WGS84.shp")
plot(reg, "DEN_REG", col = hcl.colors(length(reg), "Set 2"))
```

`data/ISTAT/Limiti01012024_g/ProvCM01012024_g/ProvCM01012024_g_WGS84.shp` contains the polygons of the provinces (administrative division between commune and region) of Italy:
```{r province-show}
italy <- vect("data/ISTAT/Limiti01012024_g/ProvCM01012024_g/ProvCM01012024_g_WGS84.shp")
plot(italy, "DEN_UTS", col = colors()[sample(seq_along(colors()), length(italy))])
```

`data/wc2.1_10m_bio_1.tif` is the mean annual temperature (averaged across several decades) at 10 minute resolution from WorldClim. I cropped this for  Europe (excluding parts of Scandinavia to save disk space):
```{r bio1-show}
bio1 <- rast("data/wc2.1_10m_bio_1.tif")
plot(bio1, col = hcl.colors(100, "Zissou 1"))
```
