# Vector operations

## Loading vectors

Load the vector shapefile of EU countries:
```{r}
eu <- vect("data/EU/EU.shp")
eu
```

This vector contains the polygons of 27 countries in the European Union (EU).
Additionally, an attribute table is attached to it, providing information about the country name, the ISO3 code, etc.
You can see full list of attributes using `names(eu)`:

```{r}
names(eu)
```

## Plotting vectors
Let's put it on a map:

```{r}
plot(eu, col = "dodgerblue")
north() #show the North
sbar(1e3, type = "bar", divs = 4, below = "kilometers", xy = c(-10, 60)) #show a scale bar
```

If you are familiar with maps, you may recognize that this is a longitude-latitude projection: the axes represent arc-degrees.
To be sure, we check the CRS of the vector:

```{r}
crs(eu, proj = TRUE)
```

As expected, this is a _longlat_ CRS.
The argument `proj = TRUE` specifies that we want a PROJ-string output.
This is more readable than the standard output of `terra`:

```{r}
crs(eu, proj = FALSE)
```

For some cases, for instance calculating areas of polygons, _longlat_ CRS are recommended.

## Area extent
In `terra`, the function `expanse()` calculate areas. 
If you read the documentation, it also specified that longlat CRS is the most accurate to calculate areas.
Let's calcualte the area of each country and add it to the attribute table of the vector:

```{r area}
A <- expanse(eu, unit = "km")
eu$area <- A
eu[, c("NAME_ENGL", "area")]
```

We can now make a map with polygons filled with colors representing the county area:

```{r}
# a gradient palette with number of colors = countries
pal <- colorRampPalette(c("grey80", "tomato3", "tomato"))(nrow(eu))
pal <- rev(pal[order(eu$area)]) #arrange colors
plot(eu, col = pal) #adding a legend to vector maps is not easy and will leave it for another chapter.
north()
sbar(1e3, type = "bar", divs = 4, below = "kilometers", xy = c(-10, 60)) 
```

## Buffers and neighbours of a polygon

Our shapefile is a good shapefile, meaning that the people who created it took care that the boundaries of countries did not overlap.
Indeed, this would be a reason for tension between countries.
However, we can still trick our way to obtain the neighbors of countries.
Imagine to enlarge the countries, i.e. to expand their borders by a small amount, let's say 1km.
This opearation is called buffering:

```{r}
b <- buffer(eu, 1e5) #100km for visualizing it
plot(eu, col = "dodgerblue")
lines(b)
```

If done on one country only, then this will overlap with its adjacent neighbors (and with non-adjacent ones if they are closer than 1km).
We can use this together with `terra::intersect()` to get its neighbors:

```{r}
neigh_name <- as.list(rep(NA, nrow(eu)))
neigh_area <- as.list(rep(NA, nrow(eu)))
names(neigh_name) <- eu$NAME_ENGL
names(neigh_area) <- eu$NAME_ENGL

# iterate over countries
for (i in seq_len(nrow(eu))) {
  b <- buffer(eu[i], 1e3)
  ans <- intersect(eu[-i], b)
  neigh_name[[i]] <- ans$NAME_ENGL
  neigh_area[[i]] <- ans$area
}
```
Some warning are thrown telling use that some countries do not have neighbors:
```{r}
names(neigh_name)[sapply(neigh_name, length) == 0]
```
These are all islands (Cyprus, Ireland, and Malta), so that is correct.

Importantly, `terra` has a special function to get adjanceny polygons, called (surprisingly) `adjacent()`.
This is the funciton you should normally use; the approach used above is for teaching purposes.
However, note that, in this case, the two approach give the same results:

```{r}
adj <- adjacent(eu)
neigh_name_correct <- as.list(rep(NA, nrow(eu)))
names(neigh_name_correct) <- eu$NAME_ENGL
for (i in seq_along(neigh_name_correct)) {
  neigh_name_correct[[i]] <- eu$NAME_ENGL[adj[adj[, 1] == i, 2]]
}

compare <- data.frame(
  buffering = unlist(neigh_name),
  correct = unlist(neigh_name_correct)
)
any(compare[, 1] != compare[, 2])
head(compare)
```

The reason they are identical is because there is no country close 1km or less to another and that is also not an adjacent neighbor.
