# Vector operations

## Loading vectors

Load the vector shapefile of EU countries:
```{r}
eu <- vect("data/EU/EU.shp")
eu
```

This vector contains the polygons of 27 countries in the European Union (EU).
Additionally, an attribute table is attached to it, providing information about the country name, the ISO3 code, etc.
You can see full list of attributes using `names(eu)`:

```{r}
names(eu)
```

## Plotting vectors
Let's put it on a map:

```{r}
plot(eu, col = "dodgerblue")
north() #show the North
sbar(1e3, type = "bar", divs = 4, below = "kilometers", xy = c(-10, 60)) #show a scale bar
```

If you are familiar with maps, you may recognize that this is a longitude-latitude projection: the axes represent arc-degrees.
To be sure, we check the CRS of the vector:

```{r}
crs(eu, proj = TRUE)
```

As expected, this is a _longlat_ CRS.
The argument `proj = TRUE` specifies that we want a PROJ-string output.
This is more readable than the standard output of `terra`:

```{r}
crs(eu, proj = FALSE)
```

For some cases, for instance calculating areas of polygons, _longlat_ CRS are recommended.

## Area extent
In `terra`, the function `expanse()` calculate areas. 
If you read the documentation, it also specified that longlat CRS is the most accurate to calculate areas.
Let's calcualte the area of each country and add it to the attribute table of the vector:

```{r}
A <- expanse(eu, unit = "km")
eu$area <- A
eu[, c("NAME_ENGL", "area")]
```

We can now make a map with polygons filled with colors representing the county area:

```{r}
# a gradient palette with number of colors = countries
pal <- colorRampPalette(c("grey80", "tomato3", "tomato"))(nrow(eu))
pal <- rev(pal[order(eu$area)]) #arrange colors
plot(eu, col = pal) #adding a legend to vector maps is not easy and will leave it for another chapter.
north()
sbar(1e3, type = "bar", divs = 4, below = "kilometers", xy = c(-10, 60)) 
```


The reason they are identical is because there is no country close 1km or less to another and that is also not an adjacent neighbor.
