# Zonal statistics

Zonal statstics are calcualted for _zones_, i.e. regions defined either by a cell value (for rasters) or by polygons (for shapefiles).
Zonal calculation are implemented in _terra_ using `zonal()`.

## Zones as a shapefile
When zones are defined by a shapefile, e.g. the polygons of EU, zonal statistics are calculated.
The syntax is `zonal(x, y, fun)`, 
where `x` is the layer with the values to calculate the statistics of,
`y` is the the layer with the zones,
and `fun` is the function of the statistic.
For example, we can calculate the average value of the annual temperature for a coutry and see how it varies with latitude:

```{r zonal}
eu <- vect("data/EU/EU.shp") #load zone layer
eu_centr <- centroids(eu) #get centroids
lat <- geom(eu_centr)[, "y"] #get latitude of centroids

bio1 <- rast("data/wc2.1_10m_bio_1.tif") #load temperature layer

bio1_mean <- zonal(bio1, eu, "mean") #zonal calculation - mean value
bio1_mean <- bio1_mean[[1]] #as vector
plot(
  lat, bio1_mean,
  pch = 20, cex = 2, 
  col = "grey50",
  xlab = "Latitude",
  ylab = "Average temperature"
)
```
In the case above `x` is a raster (it usually is), but this can also be a geometry.

By using another `fun` we can obtain other statistics.
For example, we can get the minimum and maximum values of each country:
```{r zonal-2}
bio1_min <- zonal(bio1, eu, "min") #zonal calculation
bio1_min <- bio1_min[[1]] #as vector
bio1_max <- zonal(bio1, eu, "max") #zonal calculation
bio1_max <- bio1_max[[1]] #as vector
plot(
  lat, bio1_mean,
  pch = 20, cex = 2, 
  col = "grey50",
  xlab = "Latitude",
  ylab = "Average temperature",
  ylim = c(min(bio1_min), max(bio1_max))
)
points(
  lat, bio1_min,
  pch = 6, cex = 2, 
  col = "dodgerblue3"
)
points(
  lat, bio1_max,
  pch = 2, cex = 2, 
  col = "tomato"
)
```
